# Introduction to Widgets

**Learning objectives:**

- Review alternative packages for interaction between JavaScript and R.
- Dive into using `{plotly}` for graphics.
- Expand on using `{DT}` for Data Tables.
- Understand what `{crosstalk}` *is* and how to deploy it to `{shiny}`.

`{htmlwidgets}` is a succsseor to `{rCharts}`. 

HTML Widgets can be:

- Used at the R console for data analysis just like conventional R plots.
- Embedded within R Markdown documents
- Incorporated into Shiny web applications.
- Saved as standalone web pages for ad-hoc sharing via email, Dropbox, etc.
  - (My personal favorite!!!)

There are currently 127 registered widgets on [gallery.htmlwidgets.org](http://gallery.htmlwidgets.org/)

## Plotly Package

Often, the first graphing example a new user experiences is found in [R for Data Science by Hadley Wickham and Garrett Grolemund](https://r4ds.had.co.nz/index.html). In this example, the author describes the *grammer of graphics* and the `{ggplot2}` package.

```{r, first plot}
library(ggplot2)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

Technically, there is nothing wrong with a static image and no interaction. As long as you convey your intent, your job is done. Or is it?

However, is `{ggplot}` our only graphic utilty/package/service? Certainly NO! Let's discuss [plotly](https://cran.r-project.org/web/packages/plotly/index.html).

The author comments "(plotly)...has been downloaded from CRAN 4.9 million times at the time of writing this." I wanted to double check or update the stats. Using [DataScienceMeta](http://www.datasciencemeta.com/rpackages), the site lists plotly has been downloaded 10,129,165 and is the 179th most popular downloaded (out of 18,641)!

```{r, Plotly Example}
library(plotly)

plot_ly(cars, x = ~speed, y = ~dist) %>% 
  add_markers()
```
Citation: [JavaScript for R by John Coene, Chapter 3, figure 3.1](https://book.javascript-for-r.com/widgets-intro-intro.html#fig:plotly-scatter-diagram)

- Wrap existing ggplot graphics with ggplotly, making them interactive
- Initialize the chart with `plot_ly`

Key differences between ggplot and plotly:
- ggplot uses `+` to add layers
  - this isn't horrible but doesn't lend itself to symatics and often feels like a speedbump while authoring
- plotly recognizes the `{magrittr}` pipe `%>%` operator. If using this syntax for authoring, your script text flows easier.

To achieve this layering in Plotly, we use the `add_` prefix.

```{r, Layered Plotly}
library(magrittr)
library(plotly)
plot_ly(mtcars, x = ~disp) %>% 
  add_markers(y = ~mpg, text = rownames(mtcars)) %>% 
  add_lines(y = ~fitted(loess(mpg ~ disp)))

```
Citation: [JavaScript for R by John Coene, Chapter 3, figure 3.3](https://book.javascript-for-r.com/widgets-intro-intro.html#fig:plotly-layers-diagram)

## DT Package

The `{DT}` package provides an R Interface to the jQuery JavaScript Library [DataTables](https://datatables.net/). (Not to be confused or remotely the same as Data.Tables).

Unique characteristics of DataTables versus HTML tables alone are:

 - Pagination
 - Instant Search
 - Multi-column ordering
 - Use almost any datasource
 - Easily Theme-able
 - Wide Variety of extensions
 
The following is a one-line snippet example of producing an interactivec DataTable. 
```{r, Quick example of DataTables, eval=TRUE}
library(DT)
DT::datatable(cars)
```
Citation: [JavaScript for R by John Coene, Chapter 3, figure 3.4](https://book.javascript-for-r.com/widgets-intro-intro.html#fig:plotly-dt-diagram)

## Crosstalk
Crosstalk is a an add-on to the `{htmlwidgets}` package providing cross-widget interactions. Specifically supported at this time are:

- linked brushing
- filtering

A couple limitations to using Crosstalk:

- HTML widgets must be specifically modified by their authors.
- Crosstalk currently only works for linked brushing and filtering of views that show individual data points, not aggregate or summary views.
- Crosstalk is not appropriate for large data sets.

Cross talk uses a shared R6 class to wrap existing dataframes.

```{r, Cross Talk Example, eval = TRUE}
library(crosstalk)
library(d3scatter)
shared_iris <- SharedData$new(iris)
bscols(
  d3scatter(shared_iris, ~Petal.Length, ~Petal.Width, ~Species, width="100%", height=300),
  d3scatter(shared_iris, ~Sepal.Length, ~Sepal.Width, ~Species, width="100%", height=300)
)
```
Citation: [Using Crosstalk](https://rstudio.github.io/crosstalk/using.html)

```{r, CrossTalk Filtering, eval=TRUE}
shared_mtcars <- SharedData$new(mtcars)
bscols(widths = c(3,NA,NA),
  list(
    filter_checkbox("cyl", "Cylinders", shared_mtcars, ~cyl, inline = TRUE),
    filter_slider("hp", "Horsepower", shared_mtcars, ~hp, width = "100%"),
    filter_select("auto", "Automatic", shared_mtcars, ~ifelse(am == 0, "Yes", "No"))
  ),
  d3scatter(shared_mtcars, ~wt, ~mpg, ~factor(cyl), width="100%", height=250),
  d3scatter(shared_mtcars, ~hp, ~qsec, ~factor(cyl), width="100%", height=250)
)
```
Citation: [Using Crosstalk](https://rstudio.github.io/crosstalk/using.html)


## Wrap-Up

The following [link](https://rstudio.github.io/crosstalk/widgets.html) provides additional `{crosstalk}` compatible services.

Any Questions?

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/eVmvbOE3aJQ")`

<details>
<summary> Meeting chat log </summary>

```
00:07:40	russ:	Hi everyone
00:18:14	russ:	http://gallery.htmlwidgets.org/
00:18:19	russ:	http://www.datasciencemeta.com/
00:22:15	russ:	There's a whole book about {plotly} - https://plotly-r.com/index.html
00:34:24	Arthur Shaw:	For reference, here's {reactable}: https://glin.github.io/reactable/index.html
01:04:31	russ:	https://rstudio.github.io/crosstalk/widgets.html
01:04:54	Arthur Shaw:	{reactable} working with {crosstalk}: https://glin.github.io/reactable/articles/examples.html#cross-widget-interactions
01:04:57	russ:	There is a statement at the top of the compatible widgets page: "To add a widget, please submit a PR with modifications to this page.)
```
</details>
