# Basics of Building Widgets

**Learning objectives:**

- Explore how to use JS visualisation libraries from .HTML pages
- Discuss the details (data / dependencies / context) needed by a typical JS library
- Summarise how htmlwidgets ensures these details are in place

## Read and Study

So, you have found a front-end JS library you want to use from R

Now: Learn how to use that library from JS/HTML first

- USE IT!
  - Work through a tutorial

- INSTALL IT!
  - Analyse it's dependencies / installation process

- SEE HOW OTHERS USE IT!
  - Study the examples
  - Hopefully: consistent syntax

- STUDY IT!
  - Look at the API
  
- USE IT AGAIN!
  - Code up some of your use-cases using mock data, but in JS/HTML
  - Can you improve the API for your use-case?

## Candidate Libraries

Here we work out how to use some JS visualisation libraries in their natural setting: as part of
an .html page.

The files have been rewritten slightly, to emphasise similarities / differences between the
libraries.

### HTML / DOM recap

```html
<!–– index.html - Modified from Chapter 2 ––>
<html>
  <head>
    <!--
    Nondisplayed content goes here:
    - Metadata (title / character encoding)
    - _Some_ dependency loading (stylesheet / library)
    -->
  </head>
  <body>
    <!-- Displayed content goes here -->
    <p id="content">Trying JavaScript!</p>
    <script>
      var cnt = document.getElementById("content");
      cnt.innerText = "The text has changed";
    </script>
  </body>
</html>
```

Library loading

- Book examples load libraries from .html-adjacent location:

```html
<head>
  <!-- Import library -->
  <script src="highcharts.js"></script>
</head>
```

- Here, we access the libraries from CDN (Content Delivery Network)

```html
<head>
  <!-- Import library -->
  <script src="https://code.highcharts.com/highcharts.js"></script>
</head>
```

### Plotly.js

```{r, file="examples/04/plotly.html", eval = FALSE}
```

### Highcharts.js

```{r, file="examples/04/highcharts.html", eval = FALSE}
```

### Chart.js

```{r, file="examples/04/chart.html", eval = FALSE}
```

### D3.js

- TODO

## Compare and Contrast

- Where is the plot placed?
  - 'canvas' vs 'div'

- How does the plotting library know where to put the plot?
  - `PlottingLibrary.function(x, options)`
  - Is `x` an element or the element's ID?

- How is the plot configured?
  - Is the plotData an array or object?
  - Do we call a specific function for a particular plot type, or pass an option

- From an R perspective:
  - Could we readily convert a data-frame to the rawData JSON?
  - How much of the JS API should the R user be aware of?

```html
<!-- Typical structure, based on the above -->
<html xmlns="http://www.w3.org/1999/xhtml" lang="" xml:lang="">
<head>
  <!-- Import library -->
  <script src="https://blah.blah/mycharts.min.js"></script>
</head>
<body>
  <!-- element to hold visualisation -->
  <elementType id="chart" style="width:100%;height:400px;"></elementType>
  <!-- Script to create visualisation -->
  <script>
    var rawData = {
      // JSON-like data
    }

    var plotData = {
      // all options required to make the plot
    };

    var myChart = Mycharts.someFunction(elementOrID, plotData);
  </script>
</body>
</html>
```

## How it works

- How would you write a package that calls "SuperCharts.js" if htmlwidgets didn't exist?

- Create the HTML document
- Import the external libraries
- Add HTML element to hold visualisation
- Convert R-data to JSON and embed it in the HTML
- Append JS code to transfer data/element to library

- (and seemlessly work with other R/HTML/JS/CSS tools - RMarkdown / Shiny etc)

[Link to figure](https://book.javascript-for-r.com/widgets-basics.html#widgets-basics-inner-workings)

## Summary

- We studied the problem that htmlwidgets aims to solve
- And looked at how JS libraries are used within .html pages

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
