# Basics of Building Widgets

**Learning objectives:**

- Explore how to use JS visualisation libraries from .HTML pages
- Discuss the details (data / dependencies / context) needed by a typical JS library
- Summarise how htmlwidgets ensures these details are in place

## Read and Study

So, you have found a front-end JS library you want to use from R

Now: Learn how to use that library from JS/HTML first

- USE IT!
  - Work through a tutorial

- INSTALL IT!
  - Analyse it's dependencies / installation process

- SEE HOW OTHERS USE IT!
  - Study the examples
  - Hopefully: consistent syntax

- STUDY IT!
  - Look at the API
  
- USE IT AGAIN!
  - Code up some of your use-cases using mock data, but in JS/HTML
  - Can you improve the API for your use-case?

## Candidate Libraries

Here we work out how to use some JS visualisation libraries in their natural setting: as part of
an .html page.

The files have been rewritten slightly, to emphasise similarities / differences between the
libraries.

### HTML / DOM recap

```html
<!–– index.html - Modified from Chapter 2 ––>
<html>
  <head>
    <!--
    Nondisplayed content goes here:
    - Metadata (title / character encoding)
    - _Some_ dependency loading (stylesheet / library)
    -->
  </head>
  <body>
    <!-- Displayed content goes here -->
    <p id="content">Trying JavaScript!</p>
    <script>
      var cnt = document.getElementById("content");
      cnt.innerText = "The text has changed";
    </script>
  </body>
</html>
```

Library loading

- Book examples load libraries from .html-adjacent location:

```html
<head>
  <!-- Import library -->
  <script src="highcharts.js"></script>
</head>
```

- Here, we access the libraries from CDN (Content Delivery Network)

```html
<head>
  <!-- Import library -->
  <script src="https://code.highcharts.com/highcharts.js"></script>
</head>
```

### Plotly.js

```{r, file="examples/04/plotly.html", eval = FALSE}
```

### Highcharts.js

```{r, file="examples/04/highcharts.html", eval = FALSE}
```

### Chart.js

```{r, file="examples/04/chart.html", eval = FALSE}
```

## Compare and Contrast

- Where is the plot placed?
  - 'canvas' vs 'div'

- How does the plotting library know where to put the plot?
  - `PlottingLibrary.function(x, options)`
  - Is `x` an element or the element's ID?

- How is the plot configured?
  - Is the plotData an array or object?
  - Do we call a specific function for a particular plot type, or pass an option

- From an R perspective:
  - Could we readily convert a data-frame to the rawData JSON?
  - How much of the JS API should the R user be aware of?

```html
<!-- Typical structure, based on the above -->
<html xmlns="http://www.w3.org/1999/xhtml" lang="" xml:lang="">
<head>
  <!-- Import library -->
  <script src="https://blah.blah/mycharts.min.js"></script>
</head>
<body>
  <!-- element to hold visualisation -->
  <elementType id="chart" style="width:100%;height:400px;"></elementType>
  <!-- Script to create visualisation -->
  <script>
    var rawData = {
      // JSON-like data
    }

    var plotData = {
      // all options required to make the plot
    };

    var myChart = Mycharts.someFunction(elementOrID, plotData);
  </script>
</body>
</html>
```

## How it works

- How would you write a package that calls "SuperCharts.js" if htmlwidgets didn't exist?

- Create the HTML document
- Import the external libraries
- Add HTML element to hold visualisation
- Convert R-data to JSON and embed it in the HTML
- Append JS code to transfer data/element to library

- (and seemlessly work with other R/HTML/JS/CSS tools - RMarkdown / Shiny etc)

[Link to figure](https://book.javascript-for-r.com/widgets-basics.html#widgets-basics-inner-workings)

## Summary

- We studied the problem that htmlwidgets aims to solve
- And looked at how JS libraries are used within .html pages

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
00:30:44	Ryan Metcalf:	Adding reference to new element type: https://www.w3schools.com/html/html5_canvas.asp
00:31:04	LUCIO ENRIQUE CORNEJO RAMÍREZ:	are there any advantages to adding the chart in a canvas element? Compared to Plotly, which adds the graphs in a div element
00:32:08	Ryan Metcalf:	On the surface, I presume no. However, given GPU access, the Canvas Element may utilized more optimal rendering versus a Div that would use the CPU instead.
00:32:10	Ryan Metcalf:	Just a thought.
00:34:30	Ryan Metcalf:	StackOverflow is usually not the greatest reference point...but does provide some additional content to extend to other resources: https://stackoverflow.com/questions/5882716/html5-canvas-vs-svg-vs-div
00:34:43	Ryan Metcalf:	Great question Lucio!
00:35:52	LUCIO ENRIQUE CORNEJO RAMÍREZ:	Thanks for the answer :)
00:36:37	Ryan Metcalf:	(Russ to expand to answer Lucio!!! 🙂)
00:37:07	Ryan S:	I'm familiar with html <div> elements, but not so much <canvas>.  Is <canvas> just a lesser-known html element?
00:37:56	Ryan Metcalf:	Newer supported tag. Released in 2004 by Apple.
00:38:16	Ryan S:	👍
00:38:26	Ryan Metcalf:	Again, likely not the greatest reference....extending the topic: https://medium.com/@AmyScript/introduction-to-html5-canvas-8c1bad20e855#:~:text=The%20canvas%20element%20was,to%20draw%20graphics%20utilizing%20JavaScript.
00:42:16	Arthur Shaw:	On CSS, here's the canonical meme that make's Russ' point: http://gif-central.blogspot.com/2013/07/family-guy-css.html
00:43:32	Ryan S:	LOL.... I can think of a dozen other subjects that gif applies to as well.
00:57:57	LUCIO ENRIQUE CORNEJO RAMÍREZ:	thanks!
```
</details>
