# A Realistic Widget

**Learning objectives:**

- Ability to utilize `usethis::create_package` and `htmlwidgets::scafoldWidget()` libraries
- Demonstrate the ability to modify your `/inst/htmlwidgets/<packagename>.yaml` for dependencies
- Modify both R and JS files allowing for mutual exchange of data
- Create a `widgetname_html` function for inline use of your package

We are going to use the `peity.js` plugin authored by [Ben Pickle](https://github.com/benpickles/peity) which allows for a small `<svg>` pie, donut, line, or bar chart inline with your text.

Steps to begin:

1. Run `usethis::create_package("mypeityexample")`
2. Run `htmlwidgets::scaffoldWidget("mypeityexample")`

>Note: I'm switching to a new example that won't be included in our repo!!! It is inteded you can do this on your own!

## Dependencies

To make any software utility, package, or script operate we will need dependencies. The following script was extracted from [Chapter 6 A Realistic Widget | JavaScript for R](https://book.javascript-for-r.com/widgets-realistic.html).

3. Run the following script in your new RStudio Package:

```{r, eval=FALSE}
dir.create("./inst/htmlwidgets/jquery")
dir.create("./inst/htmlwidgets/peity")

peity <- paste0(
  "https://raw.githubusercontent.com/benpickles/",
  "peity/master/jquery.peity.min.js"
)
jquery <- paste0(
  "https://code.jquery.com/jquery-3.5.1.min.js"
)

download.file(
  jquery, "./inst/htmlwidgets/jquery/jquery.min.js"
)
download.file(
  peity, "./inst/htmlwidgets/peity/jquery.peity.min.js"
)
```

The script will download these two JavaScript files to your local machine. An alternative is to use a Content Delivery Network (CDN) within your HTML file.

4. We also need to add these dependencies to your `/mypeityexample/inst/htmlwidgets/mypeityexample.ymal` file.

```{r, eval=FALSE}
dependencies:
  - name: jQuery
    version: 3.5.1
    src: htmlwidgets/jquery
    script: jquery.min.js
  - name: peity
    version: 3.3.0
    src: htmlwidgets/peity
    script: jquery.peity.min.js
```

> Question for our group - How do I know what versions are required...for ANYTHING? How can I confirm and/or trace these dependencies?

5. Validate the local copies of your dependencies are being used:

```{r, eval = FALSE}
devtools::load_all()
system.file("htmlwidgets/peity", package = "mypeityexample")
[1] "/Users/rmetcalf/Documents/GitHub/mypeityexample/inst/htmlwidgets/peity"
```

> NOTE: Since I changed the name of the package to `mypeityexample` while also following the instructions in step 3....I had to modify my system.file and my package referene above to recieve my output.

6. The final test of dependencies check is to run your package. It can be as simple as `mypeityexample("test")`. The important point here is to open your browsers DevTools, expand the <head> tag and look for the two script dependencies you included in your .yaml file.

## Implementation

## HTML Element


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/YjY8YQNh9J0")`

<details>
<summary> Meeting chat log </summary>

```
00:09:23	russ:	Hi everyone
00:11:57	russ:	Apologies for my pronunciation
00:20:13	russ:	https://www.npmjs.com/package/peity?activeTab=dependencies
00:25:18	Arthur Shaw:	Following on what Russ said, the peity library on npm shows no dependencies, but the use basic usage example uses jQuery. At the risk of being lawyerly, perhaps the library doesn't depend on jQuery, but using it may require (or most easily be done via) jQuery. npm link: https://www.npmjs.com/package/peity
01:03:30	russ:	https://github.com/JohnCoene/javascript-for-r/blob/master/code/peity/R/peity.R
01:10:14	russ:	https://github.com/ramnathv/htmlwidgets/blob/7b9c1ea3d9fbf4736d84f1fd1178fce0af29f8e3/R/htmlwidgets.R#L257
```

</details>
