# The Full Monty

**Learning objectives:**

- THESE ARE NICE TO HAVE BUT NOT ABSOLUTELY NECESSARY

## Workflow

- Write an example using vanilla HTML/JS
- Initialise an R package
- Build scaffold for widget
- Declare any dependencies
- Add JS-side code for rendering
- Work out what data / data structure / chart options should be passed from R to JS
- Add R-side code for setting up data
- Check that the visualisation looks OK
  - `<widget>_html`

## gio.js in .HTML

The following shows how to use gio.js in a plain .html file.
The code has been modified from that in the book:

- libraries are loaded from CDN
- the missing 'data' variable has been filled in

The code is very similar to the main example on the gio.js homepage.

```{r file="examples/07/gio.html", eval = FALSE}
```

## R package

We create an R package with a widget.

```r
usethis::create_package("giro")
htmlwidgets::scaffoldWidget("newGio")
```

We've done this a few times now:

- `newGio.[R|yaml|js]` files are created and opened

## Declaring Dependencies

We use v2.0.0 of gio.js and v110 of three.js, as in the book

We add a ./dev script to contain the code that downloaded the dependencies:

```r
# ./dev/download-dependencies.R

versions <- list(
  three = "110",
  gio = "2.0.0"
)

dirs <- list(
  htmlwidgets = file.path("inst", "htmlwidgets")
)
dirs$three <- file.path(dirs$htmlwidgets, "three")
dirs$gio <- file.path(dirs$htmlwidgets, "gio")

urls <- list(
  three = glue::glue(
    "https://cdnjs.cloudflare.com/ajax/libs/three.js/{versions$three}/three.min.js"
  ),
  gio = glue::glue(
    "https://cdn.jsdelivr.net/npm/giojs@{versions$gio}/build/gio.min.js"
  )
)

# create directory for JS dependencies
dir.create(dirs$three, recursive = TRUE)
dir.create(dirs$gio, recursive = TRUE)

# download the JS dependencies
download.file(urls$three, file.path(dirs$three, "three.min.js"))
download.file(urls$gio, file.path(dirs$gio, "gio.min.js"))
```

We have done this before:

- Generate an R package
- with the scaffold for a widget
- and the dependencies of that widget downloaded into subdirectories.

Now

- Tell {htmlwidgets} where the libraries can be found:

```yaml
dependencies:
  - name: three
    version: 110
    src: htmlwidgets/three
    script: three.min.js
  - name: gio
    version: 2.0
    src: htmlwidgets/gio
    script: gio.min.js
```

Put "three.js" first, since "gio.js" depends on "three.js"

## JS-side code

Skeleton for adding a GIO visualisation is placed into 'renderValue' function in the *.js file:

Note that `el` is the element within which the GIO visualisation will be placed

```js
// ... snip ...
factory: function(el, width, height) {
  return {
    renderValue: function(x) {
      // Iterations:
      // var container = document.getElementById("globe"); // hard-coded element ID
      // var container = document.getElementById(el.id); // use the ID for the user's element
      // var controller = GIO.Controller(container);
      // ... but `container` _is_ `el`
      
      var controller = GIO.Controller(el);
      // controller.addData(data);
      controller.init();
    }
  }
  // ... snip ...
}
```

With this we can generate an empty visualisation and add it to the user-supplied element `el`.

## Working with data

The data expected by GIO is as follows:

```js
// Array[Object{e: ISO2CODE, i: ISO2CODE, v: number}]
[
  {
    "e": "CN", // exporting country
    "i": "US", // importing country
    "v": 3300000 // value
  },
  {
    "e": "CN",
    "i": "RU",
    "v": 10000
  }
]
```

https://giojs.org/html/docs/dataAdd.html

R reads that as a data-frame:

```{r}
arcs <- jsonlite::fromJSON(r"([
  {
    "e": "CN", // exporting country
    "i": "US", // importing country
    "v": 3300000 // value
  },
  {
    "e": "CN",
    "i": "RU",
    "v": 10000
  }
])")
```

```{r}
arcs
```

Conflict between R-style datasets and GIO-expected / JS-style datasets

- Objects in the JS array ~ Rows of the R data-frame

Widget changes:
- use `data` param rather than `message` in R function
- include this as `x.data` in the object passed to JS
- Use `x.data` in the `controller.addData()` call

```js
// ... snip ...
factory: function(el, width, height) {
  return {
    renderValue: function(x) {
      var controller = GIO.Controller(el);
      controller.addData(x.data);
      controller.init();
    }
  }
  // ... snip ...
}
```

```r
newGio <- function(data, width = NULL, height = NULL, elementId = NULL) {

  # forward options using x
  x = list(
    data = data
  )
# ... snip ...
```

Look in devtools at the contents of the dataset:

```js
<script type="application/json" data-for="htmlwidget-b97f43c69fa089f58c24">
{
  "x": {
    "data":{
      "e":["CN","CN"],
      "i":["US","RU"],
      "v":[3300000,10000]
    }
  },
  "evals":[],
  "jsHooks":[]
}
</script>
```

Columns in R have become the elements in `x.data`

## Transforming Data

We want the serialised data to be in the format the JS library expects

Approaches:

- Convert the data on the JS side
  - `HTMLWidgets.dataframeToD3()`
- Modify the data in R before serialising it
  - e.g., make list of rows
- Modify the serialiser used by htmlwidgets
- Replace the serialiser used by htmlwidgets

The default serialiser in htmlwidgets uses `jsonlite::toJSON(..., dataframe = "columns")`:

```{r convert_columns_to_json}
# This is not what we want ...
jsonlite::toJSON(arcs, dataframe = "columns")
```

```{r convert_rows_to_json}
# This is what we want ...
jsonlite::toJSON(arcs, dataframe = "rows")
```

```{r jsonify_example}
jsonify::to_json(arcs, unbox = TRUE)
```

To alter the serialiser used by {htmlwidgets} you modify the attributes of the `x` that you pass
into `createWidget`.

Either with "TOJSON_ARGS" or "TOJSON_FUNC".

```{r modifying_the_default_serialiser}
# modify the arguments of the default serialiser:

# ... snip
x <- list(data = arcs)
attr(x, "TOJSON_ARGS") <- list(dataframe = "rows")
```

```{r replacing_the_default_serialiser}
# replace the default serialiser
gio_serialiser = function(x) {
  jsonify::to_json(x, unbox = TRUE)
}

# ... snip
x = list(data = arcs)
attr(x, "TOJSON_FUNC") <- gio_serialiser
```

See the Data Transformation section of one of the {htmlwidgets} vignettes: https://www.htmlwidgets.org/develop_advanced.html#data-transformation

## On Print Method

In ggplot2:

```{r two_steps_of_ggplot2}
library(ggplot2)

# create an object
my_plot <- ggplot(diamonds, aes(x = x, y = price)) + geom_point()

# plot the object
my_plot
```

With a widget:

```r
# create an object
my_widget <- gio(arcs)

# visualise the object
my_widget
```

The object created can be subsequently modified before visualisation ('print'ing).

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
