# Linking Widgets

**Learning objectives**:

- Learn how `crosstalk` links widgets
- Explore linking widgets with gio as an example

## Motivating example

### Basic

```{r crosstalk_basic, eval=FALSE}
library(DT)
library(plotly)
library(crosstalk)

shared <- SharedData$new(cars)

bscols(
  plot_ly(shared, x = ~speed, y=~dist),
  datatable(shared, width = "100%")
)
```

### Shiny

```{r crosstalk_shiny, eval=FALSE}
library(DT)
library(shiny)
library(crosstalk)

ui <- fluidPage(
  fluidRow(
    column(4, uiOutput("text")),
    column(8, DTOutput("table"))
  )
)

server <- function(input, output) {
  sd <- SharedData$new(cars)

  output$text <- renderUI({
    # get selected rows
    n_selected <- sd$data(withSelection = TRUE) %>% 
      dplyr::filter(selected_ == TRUE) %>% 
      nrow()

    h3(n_selected, "selected items")
    
  })

  output$table <- renderDT({
    datatable(sd)
  }, server = FALSE)
}

shinyApp(ui, server)
```

## How `crosstalk` "talks"

### Shared data

```{r shared_data, eval=TRUE}
# create shared data...
sd_cars <- SharedData$new(cars[1:2,])

# ...that is not a simple data frame
class(sd_cars)
str(sd_cars)
```

### Communication 

#### Keys

Keys are the key for communication

- Action in widget 1
  - Listen for actions
  - Capture key where action occurs
  - Pass key to widget 2
- React in widget 2
  - Listen for keys emitted by widget 1
  - Receive key from widget 1
  - Make changes in widget 2 using key from widget 1

Where do keys come from?

- Either bring them--for example, `rownames` in a dataframe
- Or set them (in the package logic)

#### Groups

- Keys identify observations in the same group
- Groups are
  - Either for the same data set
```{r, eval=FALSE}
# same data; shared keys, by construction
shared_cars <- SharedData$new(mtcars, group = "cars")
```
  - Or for different data sets with shared keys
```{r, eval=FALSE}
# different data; shared keys
shared_cars       <- SharedData$new(mtcars, group = "cars")
shared_cars_head  <- SharedData$new(head(mtcars), group = "cars")
```

## Crosswalk with `gio`

### R code

What to change:

- Accept shared data frame, rather than simple data frame: `if (crosstalk::is.SharedData(data))`
- Extract original data from shared data object: `data <- data$origData()` 
- Extract group: `group <- data$groupName()`
- Set `crosstalk` dependency:
  - `deps <- crosstalk::crosstalkLibs()`
  - `htmlwidgets::createWidget(dependencies = deps)`

```{r, eval=FALSE}
gio <- function(data, width = NULL, height = NULL, 
  elementId = NULL) {

  # defaults to NULL
  group <- NULL
  deps <- NULL

  # uses crosstalk
  if (crosstalk::is.SharedData(data)) {
    group <- data$groupName()
    data <- data$origData()
    deps <- crosstalk::crosstalkLibs()
  }

  # forward options using x
  x = list(
    data = data,
    style = "default",
    crosstalk = list(group = group) # pass group
  )

  attr(x, 'TOJSON_ARGS') <- list(dataframe = "rows")

  # create widget
  htmlwidgets::createWidget(
    name = 'gio',
    x,
    width = width,
    height = height,
    package = 'gio',
    elementId = elementId,
    sizingPolicy = htmlwidgets::sizingPolicy(
      padding = 0,
      browser.fill = TRUE,
      defaultWidth = "100%"
    ),
    preRenderHook = render_gio,
    # add crosstalk dependency
    dependencies = deps
  )
}
```

### JS code

#### Register selections

What to change:

- Create selection handler: `var sel_handle = new crosstalk.SelectionHandle();`
- Set the group passed from R: `sel_handle.setGroup(x.crosstalk.group);`

#### Send selection from `gio` to another widget

```{js, eval=FALSE}
function callback (selectedCountry) {
  sel_handle.set([selectedCountry.ISOCode]);
}

controller.onCountryPicked(callback);
```

#### Receive selection from another widget; react in  `gio`

```{js, eval=FALSE}
// placed in factory function
sel_handle.on("change", function(e) {

  // selection comes from another widget
  if (e.sender !== sel_handle) {
    // clear the selection
    // not possible with gio.js
  }
  controller.switchCountry(e.value[0]);
});
```

### Usage example

```{r gio_w_crosstalk, eval=FALSE}
library(DT)
library(gio)
library(crosstalk)

url <- paste0(
  "https://raw.githubusercontent.com/JohnCoene/",
  "javascript-for-r/master/data/countries.json"
)
arcs <- jsonlite::fromJSON(url)

# Wrap data frame in SharedData
# key is importing country
sd <- SharedData$new(arcs, key = arcs$i)

bscols(
  gio(sd),
  datatable(sd, width="100%", selection = "single")
)
```

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/8mgCUBGMzEc")`

<details>
<summary> Meeting chat log </summary>

```
00:01:04	LUCIO ENRIQUE CORNEJO RAM REZ:	Hello
00:01:15	russ:	Hi everyone
00:04:14	Ryan Metcalf:	Forgive my delayed joining.
00:23:40	Ryan Metcalf:	https://r6.r-lib.org/articles/Introduction.html
01:08:31	LUCIO ENRIQUE CORNEJO RAM REZ:	thanks!!
```
</details>
