# Bidirectional Communication

**Learning objectives:**

- Explore Shiny's ability to communicate the R server and the user interface.
- Lean how to include JavaScript in Shiny applications.

## WebSocket an Shiny

- A **WebSocket** is a persistent connection which will allow us
to communicate Shiny's main components:
    - **user interface** (`ui`)
    - **server function** (`server`)

```{r, websocket-diag}
#| echo: false 
#| fig.pos: "H"
#| fig.align: center
#| out.width: "100%"
#| fig.dim: c(3,6)
#| fig.cap: 'Websocket visualised'
DiagrammeR::grViz("
digraph{
  node [shape=record fontsize=10];
  w1 [label=Websocket];
  w2 [label=Websocket];
  w3 [label=Websocket];
  w4 [label=Websocket];
  b1 [label='Browser' color=gold];
  b2 [label='Browser' color=gold];
  b3 [label='Browser' color=gold];
  b4 [label='Browser' color=gold];
  s [label='R server' color=royalBlue];
  s -> w1 -> b1 [dir=both color=dimGray arrowsize=.3];
  s -> w2 -> b2 [dir=both color=dimGray arrowsize=.3];
  s -> w3 -> b3 [dir=both color=dimGray arrowsize=.3];
  s -> w4 -> b4 [dir=both color=dimGray arrowsize=.3];
}
")
```

Basic example of bidirectional communication:
  
- Send text input value to the R server, 
and then send it back to the UI.

```r
library(shiny)

ui <- fluidPage(
  textInput("nameInput", "Your name"),
  textOutput("nameOutput")
)

server <- function(input, output) {
  output$nameOutput <- renderText({
    input$nameInput
  })
}

shinyApp(ui, server)
```

- WebSocket in action

```{r, shiny-websocket-diagram}
#| fig.pos: "H"
#| echo: false
#| out.width: "100%"
#| fig.dim: c(3, 6)
#| fig.align: center
#| fig.cap: 'Shiny websocket visualised'
DiagrammeR::grViz("
digraph {
  graph[rankdir=LR fontsize = 10]
  node[shape=record fontsize = 9]
  subgraph cluster_0 {
    textInput
    textOutput
    label='User Interface'
    color=gold
  }
  subgraph cluster_1 {
    'input list'
    renderOutput
    label='R server'
    color=royalBlue
  }
  textInput -> 'input list' [xlabel=websocket fontsize = 8]
  'input list' -> renderOutput
  renderOutput -> textOutput [label=websocket fontsize = 8]
}
")
```

JavaScript is required for the communication between the
R server and the user interface.

## Sessions

As seen in \@ref(fig:websocket-diag), for any Shiny application,
Shiny isolates its users into what it referes to as a **session**.

This way, any valid user interaction with a Shiny application
updates only what is being shown in his or her screen.

## Alerts, an example

Goals:

- Build a Shiny application using the **jBox** JavaScript library.
- Such application will display a notification at the click of an
`actionButton` and will *tell* the server when such notification is closed.

```js
// Type into the browser console
alert("Alert message")
```

### Explore

Let's load **jBox** into a Shiny app, in order to play around with its features.

```r
library(shiny)

ui <- fluidPage(
  tags$head(
    tags$script(
    # Load jBox's script
      src = paste0(
        "https://cdn.jsdelivr.net/gh/StephanWagner/",
        "jBox@v1.2.0/dist/jBox.all.min.js"
      )
    ),
    # Load jBox's styling
    tags$link(
      rel = "stylesheet",
      href = paste0(
        "https://cdn.jsdelivr.net/gh/StephanWagner/",
        "jBox@v1.2.0/dist/jBox.all.min.css"
      )
    )
  ),
  # Create a jBox alert
  tags$script("
    new jBox('Notice', {
      content: 'Hurray! A notice!'
    });"
  ),
  # Add a title and set its HTML element identifier 
  h1("Exploring jBox", id = "title")
)

server <- function(input, output) {}

shinyApp(ui, server)
```

Be careful when searching for **jBox**'s 
[documentation](https://stephanwagner.me/jBox/documentation),
do not Google **jBox**; type **jBox js** instead.

A couple more **jBox** features:

```js
// Type into the browser console
new jBox(
  // Set plugin: Tooltip, Modal, Notice, ... 
  'Tooltip', 
  {
    attach: '#title',
    trigger: 'click',
    onCreated: function () {
      this.setContent('This is my jBox!');
    },
    onOpen: function () {
      new jBox('Notice', {content: 'Here I come!', color: 'green'});
    },
    onClose: function () {
      new jBox('Notice', {content: 'See ya later!', color: 'red'});
    }
  }
);
```

### From R to JavaScript

Goal: Dynamically set the message in the alert created with **jBox**.

Data can be sent (through the WebSocket) from the server to the client,  
via **R**'s `session` object and its `sendCustomMessage` method/function.

Let's update the `server` function of our previous app

```r
server <- function(input, output, session) {
  session$sendCustomMessage(
    # Set the identifier to send-notice,
    # for JavaScript to "catch"
    type = "send-notice", 
    # Data to send to JavaScript
    message = "Hi there!"
  )
}
```

JavaScript's `Shiny` object and its method `addCustomHandler`
provides to JavaScript the instructions for when data is sent
from the server, with a certain *type*/*identifier*.

```js
tags$script(
  "Shiny.addCustomMessageHandler(
    type = 'send-notice', 
    function (message) {
      new jBox('Notice', {
        content: message
      });
    }
  );"
)
```

Similar example:

```r
library(shiny)

ui <- fluidPage(
  tags$head(
    tags$script(
      src = paste0(
        "https://cdn.jsdelivr.net/gh/StephanWagner/",
        "jBox@v1.2.0/dist/jBox.all.min.js"
      )
    ),
    tags$link(
      rel = "stylesheet",
      href = paste0(
        "https://cdn.jsdelivr.net/gh/StephanWagner/",
        "jBox@v1.2.0/dist/jBox.all.min.css"
      )
    )
  ),
  textInput("msg", "notice"),
  actionButton("notify", "Notify myself"),
  tags$script(
    "Shiny.addCustomMessageHandler(
      type = 'send-notice', 
      function (message) {
        new jBox('Notice', {
          content: message
        });
        // How was the message data converted?
        console.log(message, typeof message);
      }
    );"
  )
)

server <- function(input, output, session) {
  observeEvent(input$notify, {
    session$sendCustomMessage(
      type = "send-notice",
      message = paste0(input$msg, "!!!")
    )
  })
}

shinyApp(ui, server)
```

```{r, shiny-alert-diagram}
#| fig.pos: "H"
#| echo: false
#| fig.align: 'center'
#| out.width: "100%"
#| fig.dim: c(3, 6)
#| fig.cap: 'Shiny alert with custom messages'
DiagrammeR::grViz("
digraph{
  graph [rankdir=LR]
  node [shape=record fontsize=10];
  subgraph cluster_0 {
    textInput
    handler [label='message handler']
    notice [label = 'show notice']
    label='Client'
    color=gold
  }
  subgraph cluster_1 {
    send[label='sendCustomMessage']
    label='Server'
    color=royalBlue
  }
  textInput -> send [label='value'];
  send -> handler [label='message'];
  handler -> notice;
}
")
```

### Serialisation

- Note that the **jBox** notice is configured using a JSON object:

```js
new jBox('Notice', {
  content: 'Hurray! A notice!',
  "color": 'red'
});
```

- **R**'s analogue for JSON object:

```{r}
#| comment: ""
options <- list(
  content = 'Hurray! A notice!',
  color = 'red'
)
jsonlite::toJSON(options, pretty = TRUE, auto_unbox = TRUE)
```

- Using **R**'s lists for sending data in JSON format:

```r
library(shiny)

ui <- fluidPage(
  tags$head(
    tags$script(
      src = paste0(
        "https://cdn.jsdelivr.net/gh/StephanWagner/",
        "jBox@v1.2.0/dist/jBox.all.min.js"
      )
    ),
    tags$link(
      rel = "stylesheet",
      href = paste0(
        "https://cdn.jsdelivr.net/gh/StephanWagner/",
        "jBox@v1.2.0/dist/jBox.all.min.css"
      )
    ),
    tags$script(
      "Shiny.addCustomMessageHandler(
        type = 'send-notice',
        function (message) {
          // Use notice sent from the server
          new jBox('Notice', message);
        }
      );"
    )
  )
)

server <- function(input, output, session){
  # Define notice options
  notice = list(
    content = 'Hello from the server',
    color = 'black'
  )
  # Send the notice
  session$sendCustomMessage(
    type = "send-notice",
    message = notice
  )
}

shinyApp(ui, server)
```

### JavaScript to R

**Goal:** Send data from JavaScript to R when the notice is closed.

Remember that such event (notice gets closed) can be used like this:

```js
// Type into the browser console
new jBox(
  // Set plugin: Tooltip, Modal, Notice, ... 
  'Tooltip', 
  {
    attach: '#title',
    trigger: 'click',
    onCreated: function () {
      this.setContent('This is my jBox!');
    },
    onOpen: function () {
      new jBox('Notice', {content: 'Here I come!', color: 'green'});
    },
    onClose: function () {
      new jBox('Notice', {content: 'See ya later!', color: 'red'});
    }
  }
);
```

- We'll send data from JavaScript to the R Server, via 
a **simplified Shiny input**.

```js
Shiny.setInputValue('inputId', inputValue)
```

- Example app different than the one provided in the main book:

```r
library(shiny)

ui <- fluidPage(
  tags$head(
    # Load jBox dependencies
    tags$script(
      src = paste0(
        "https://cdn.jsdelivr.net/gh/StephanWagner/",
        "jBox@v1.2.0/dist/jBox.all.min.js"
      )
    ),
    tags$link(
      rel = "stylesheet",
      href = paste0(
        "https://cdn.jsdelivr.net/gh/StephanWagner/",
        "jBox@v1.2.0/dist/jBox.all.min.css"
      )
    ),
    # Set custom message handler
    tags$script(
      "Shiny.addCustomMessageHandler(
        type = 'send-notice', function (message) {
          message.onClose = function (){
            Shiny.setInputValue(
              'notice_close', 
              document.getElementById('time').innerText
            );
          }
          new jBox('Notice', message);
      });"
    )
  ),
  tags$body(
    # Show time counter
    div(
      span("Time counter (in seconds): ", style = "font-weight: bold"),
      span("", id = "time")
    ),
    br(),
    # Show notice
    actionButton("show", "Show the notice"),
    # Update the time after each second
    tags$script(
      "let time = 0;
      setInterval(
        // Update second counter
        function () {
          time += 1;
          document.getElementById('time').innerText = time;
        },
        // Repeat every second
        1000
      );
    "),
  )
)

server <- function(input, output, session){
  observeEvent(input$show, {
    # Define notice options
    notice = list(
      content = "Click to send data to the R Server",
      color = 'black'
    )
    # Send the notice
    session$sendCustomMessage(
      type = "send-notice", message = notice
    )
  })

  # Print the output of the notice_close event (when fired)
  observeEvent(input$notice_close, {
    print(input$notice_close)
  })
}

shinyApp(ui, server)
```

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
