# Custom Outputs

**Learning objectives:**

- Define custom `render*()` and `*Output()` functions
- Describe the purpose of `render*()` and `*Output()` functions
- Explain "output binding"s in JavaScript
- Explain how to conditionally / dynamically load dependencies

## Setup

- In the book they make an app and add functions `boxxy`, `renderBoxxy` and `boxxyOutput` to the
  R/ directory
- Here, we'll make a package containing the `boxxy` functions and add an app to demonstrate it

```r
library(usethis)
create_package("boxxy")
use_package("shiny")
use_mit_license()
```

```bash
# BASH
# See https://www.npmjs.com/package/gitignore
gitignore r
```

```r
# Copied a function from Dean attali's blog to run a named app from
# ./inst/shiny-examples/app-name/
# - See https://deanattali.com/2015/04/21/r-package-shiny-app/
# - called it run_example(app_name)
usethis::use_r("run_example")
```

```bash
# BASH
# Added a shiny app to ./inst/shiny-examples/boxxy-demo
mkdir -p ./inst/shiny-examples/boxxy-demo
touch ./inst/shiny-examples/boxxy-demo/app.R
```

```bash
# BASH
# The dependencies of `boxxy` will be put into `./inst/assets`
mkdir -p ./inst/assets
```

```r
# I'd probably put this in ./dev/setup_dependencies.R
dirs <- list(
  assets = file.path("inst", "assets")
)

urls <- list(
  countup = paste0(
    "https://cdn.jsdelivr.net/npm/countup@1.8.2/countUp.js"
  )
)

paths <- list(
  countup = file.path(dirs$assets, "countup.js"),
  binding = file.path(dirs$assets, "binding.js"),
  css = file.path(dirs$assets, "styles.css")
)

# Download countUp.js
download.file(urls$countup, paths$countup)

# create  binding file
file.create(paths$binding)

# create CSS file
file.create(paths$css)
```

So our package outline looks like this (before we've added the `boxxy()` functions).

```bash
tree .
.
├── boxxy.Rproj
├── DESCRIPTION
├── inst
│   ├── assets
│   │   ├── binding.js
│   │   ├── countup.js
│   │   └── styles.css
│   └── shiny-examples
│       └── boxxy-demo
│           └── app.R
├── LICENSE
├── LICENSE.md
├── man
│   └── run_example.Rd
├── NAMESPACE
└── R
    └── run_example.R

6 directories, 11 files
```

## Output R functions

We are making an animated box, containing a count, for which the user can specify a title, value
and color.

We will write the following functions:

- `boxxy()`
  - stores the data for generating the box (creates a boxxy object)
- `boxxyOutput()`
  - defines the HTML to be added to the app for the boxxy element
- `renderBoxxy()`
  - creates R logic that executes when a matching boxxy output is requested by the browser

We will also have to write a JS binding function to define how a rendered boxxy element should be
displayed inside the app.

If you've used shiny before, you should be used to the naming conventions.

The functions will be called like this:

```r
# In the shiny UI:
boxxyOutput("boxxyID")

# ... snip ...

# In the shiny server:
output$boxxyID <- renderBoxxy({
  boxxy(title, value, color)
})
```

```r
usethis::use_r("boxxy")
```

The skeleton for the R functions:

```r
#' Stores data for creating an animated, counted, box
#'
#' @param   title,value,color   title, value and color for the box
#' @export

boxxy <- function(title, value, color = "#ef476f") {
  list(title = title, value = value, color = color)
}

#' Create the HTML skeleton within which a boxxy element sits
#'
#' @param   id   A shiny identifier
#' @export

boxxyOutput <- function(id) {
  shiny::tags$div(id = id, class = "boxxy")
}

#' Render a 'boxxy' object for output in a shiny app
#'
#' @param   expr   a shiny expression for creating a 'boxxy' object.
#' @param   env   the environment in which to evaluate \code{expr}.
#' @export

renderBoxxy <- function(expr, env = parent.frame()) {
  function() {}
}
```

## Generate Output HTML

`boxxyOutput(id)`

- returns an HTML tag which
- ... should have an `id` or a `data-input-id` attribute
- ... and a `class` 

```{r output_generated_html, results="asis"}
# The ID passed into a *Output is the ID in the created element
# - Note that the parameter is `outputId` not `id` (as in the book)
message(
  shiny::plotOutput(outputId = "my_plot")
)
```

JS uses the `class` to find elements in the DOM and to generate output.

The simplest output function:

```r
#' Create the HTML skeleton within which a boxxy element sits
#'
#' @param   outputId   A shiny identifier
#' @export

boxxyOutput <- function(outputId) {
  shiny::tags$div(id = outputId, class = "boxxy")
}
```

But, the boxes are supposed to have a title and value (and color - see later):

```{r boxxy_output_with_value_and_title}
# ./R/boxxy.R
boxxyOutput <- function(outputId) {
  shiny::tags$div(
    id = outputId, class = "boxxy",
    shiny::h1(
      id = sprintf("%s-boxxy-value", outputId),
      class = "boxxy-value"
    ),
    shiny::p(
      id = sprintf("%s-boxxy-title", outputId),
      class = "boxxy-title"
    )
  )
}
```

```{r boxxy_output_example}
message(
  boxxyOutput("my_box")
)
```

The `class`es can be used within CSS to help with styling

```css
/* ./inst/assets/styles.css */
.boxxy {
  text-align: center;
  border-left: 6px solid #073b4c;
  padding: 1em;
}

.boxxy-title {
  text-transform: uppercase;
}

.boxxy-value {
  font-size: 3em;
}
```

Dependencies:

- in {boxxy}, we have assets/{countup,binding}.js and assets/styles.css
- these need to be attached to the DOM if boxxyOutput() is called by a shiny app
- so let's attach them when `boxxyOutput()` is called ...

```{r boxxy_output_with_deps}
boxxyOutput <- function(outputId) {
  value_el <- shiny::h1(
    id = sprintf("%s-boxxy-value", outputId), class = "boxxy-value"
  )
  title_el <- shiny::p(
    id = sprintf("%s-boxxy-title", outputId), class = "boxxy-title"
  )
  
  boxxy_el <- shiny::tags$div(id = outputId, class = "boxxy", value_el, title_el)
  
  deps <- list(
    htmltools::htmlDependency(
      name = "boxxy",
      version = "1.0.0",
      src = c(file = "assets"),
      script = c("countup.js", "binding.js"),
      stylesheet = "styles.css",
      package = "boxxy"
    )
  )
  
  htmltools::attachDependencies(boxxy_el, deps)
}
```

Here, {boxxy} is a package not an app, so some of the code is different to the book's version.
We take the `htmlDependency` from the package `boxxy`, rather than using a normalised path.
The `src` directory needs to be indicated relative to the installed package (`assets`, rather than
`inst/assets`).

Note that the dependencies aren't seen when `boxxyOutput()` is called.

```{r boxxy_output_doesnt_display_dependencies}
message(
  boxxyOutput("a_fantastic_box")
)
```

## Output R function

```{r define_boxxy}
boxxy <- function(title, value, color = "#ef476f") {
  list(title = title, value = value, color = color)
}
```

## Output renderer

Shiny `render*()` functions accept an expression:

```r
output$myPlot <- renderPlot({
  cars %>% head() %>% plot()
})
```

... and they return a function:

```{r class_returned_by_renderX}
class(
  shiny::renderPlot({
    cars %>% head() %>% plot()
  })
)
```

shiny has a function for converting expressions (and their environments) to a function:

```{r expr_to_function}
shiny::exprToFunction({
  cars %>% head() %>% plot()
})
```

So we can define a render function for `boxxy` like this:

```{r renderBoxxy1}
renderBoxxy <- function(expr, env = parent.frame(), quoted = FALSE) {
  func <- shiny::exprToFunction(expr, env, quoted)
  
  function() {
    func()
  }
}
```

But isn't that equivalent to:

```{r renderBoxxy2}
renderBoxxy <- function(expr, env = parent.frame(), quoted = FALSE) {
  func <- shiny::exprToFunction(expr, env, quoted)
  func  
}
```

Looking at the source code for a few render functions shows quite a bit of variability:

- renderXXX as templated by htmlwidgets
  - https://github.com/jbkunst/highcharter/blob/main/R/highcharter.R#L113
- `htmlwidgets::shinyRenderWidget` (which powers renderXXX for htmlwidgets)
  - https://github.com/ramnathv/htmlwidgets/blob/400cf1ad78cf106950525d3b70971c28029e2288/R/htmlwidgets.R#L550
- `shiny::renderPlot`
  - https://github.com/rstudio/shiny/blob/d4527cdc289c897b92a0b06eef4200537c0d98b1/R/render-plot.R#L60

Why is createRenderFunction not used here?

- Docs: https://shiny.rstudio.com/reference/shiny/1.6.0/createRenderFunction.html

## JavaScript Output binding

What is an "Output binding"?

- JS object
- Tells shiny how to find the output component
- ... and how to interact with it

See the RStudio [article on building outputs](https://shiny.rstudio.com/articles/building-outputs.html)

    (Note that each instance of the output component doesn’t need
    its own output binding object; rather, all instances of a
    particular type of output component share a single output
    binding object.)
    
    - RStudio article

```js
// ./inst/assets/binding.js
const boxxyBinding = new Shiny.OutputBinding();
```

Code [definition](https://github.com/rstudio/shiny/blob/a8c14dab9623c984a66fcd4824d8d448afb151e7/srcts/src/bindings/output/outputBinding.ts) for `Shiny.OutputBinding`

Default methods:

- `find` (find all elements of the same class as the required output)
- `renderValue`
- `getId`
- `onValueChange`
- `onValueError`
- `renderError`
- `clearError`
- `showProgress`

Of these, you _must_ define `find()`

Adding a method to an existing target-object using JQuery's `$.extend(target, object)`

```js
// ./inst/assets/binding.js
const boxxyBinding = new Shiny.OutputBinding();

$.extend(boxxyBinding, {
  find: function(scope) {
    // JQuery to get an array of elements of class "boxxy"
    return $(scope).find(".boxxy");
  }
});
```

You also need a way to get the HTML ID for a given element

```js
// getId default
this.getId = function(el) {
  return el["data-input-id"] || el.id;
}
```

You also need to define a `renderValue(el, data)` method which converts the data passed by the
R-side `renderBoxxy()` function and attaches it to the element created by `boxxyOutput()`

And, register the output binding with shiny.

Difficulties here:

- For boxxy, `find` returns an array of elements of class `boxxy`
- `getId` returns the ID of a given one of those elements
- So when I change the data passed to a boxxy element in shiny
- ... how does it know which of the DOM's boxxy elements to update?

For each registered OutputBinding:

- JS-side shiny iterates over all elements that are returned by `.find()`
- .. and makes an adapter that connects the element to it's OutputBinding
- see the [bind.ts source code](https://github.com/rstudio/shiny/blob/a8c14dab9623c984a66fcd4824d8d448afb151e7/srcts/src/shiny/bind.ts#L122)

We now fill in the `renderValue` method (to tell JS what to do when new data is received)

Remember what `boxxyOutput()` returns:

```{r calling_boxxy_output}
message(
  boxxyOutput("my_id")
)
```

Remember what `boxxy()` receives and outputs:

```{r calling_boxxy}
boxxy(title = "Some title", value = 87, color = "#FF00FF")
```

The 'title'  needs to be presented by the `<p id="<the_id>-boxxy_title" class="boxxy-title">`
element.

```js
// ./inst/assets/binding.js
const boxxyBinding = new Shiny.OutputBinding();

$.extend(boxxyBinding, {
  // ... snip ...
  renderValue: function(el, data) {
    // insert the title
    const titleId = el.id + "-boxxy-title";
    document.getElementById(titleId).innerText = data.title;
  }
});
```

Similarly, we can set the counter value and the background color for the element:

```js
// ./inst/assets/binding.js
const boxxyBinding = new Shiny.OutputBinding();

$.extend(boxxyBinding, {
  // ... snip ...
  renderValue: function(el, data) {
    // insert the title
    const titleId = el.id + "-boxxy-title";
    document.getElementById(titleId).innerText = data.title;
    
    // place counter in elementId
    // start at 0 and count up to 123
    const counterId = el.id + "-boxxy-value";
    const counter = new CountUp(counterId, 0, data.value);
    counter.start();
    
    // background color
    el.style.backgroundColor = data.color;
  }
});
```

To register an OutputBinding with shiny, we use the following code:

```js
// ./inst/assets/binding.js
//
// The second argument doesn't do anything yet
Shiny.outputBindings.register(boxxyBinding, "john.boxxy");
```

## Boxxy Usage

In the book a standalone app was made (`./app.R`, `./assets/*`).

Here, we made a package `{boxxy}` that contains:

- the `boxxy`, `renderBoxxy`, `boxxyOutput` R functions
- the external countup.js dependency (`./inst/assets/countup.js`)
- some css styling (`./inst/assets/styles.css`)
- a custom binding.js file that tells JS-Shiny how to update any boxxy output elements 
(`./inst/assets/binding.js`)

Since this is a package, our demo app looks simpler than that in the book:

```r
library(shiny)
library(boxxy)

ui <- function() {
  fluidPage(
    h2("Custom outputs"),
    fluidRow(
      column(3, boxxyOutput("countries")),
      column(3, boxxyOutput("employees")),
      column(3, boxxyOutput("customers")),
      column(3, boxxyOutput("subs"))
    )
  )
}

server <- function(input, output, session) {
  output$countries <- renderBoxxy({
    boxxy("Countries", 95, color = "#ef476f")
  })

  output$employees <- renderBoxxy({
    boxxy("Thing", 650, color = "#06d6a0")
  })

  output$customers <- renderBoxxy({
    boxxy("Customers", 13592, color = "#118ab2")
  })

  output$subs <- renderBoxxy({
    boxxy("Subscriptions", 16719, color = "#ffd166")
  })
}

shinyApp(ui(), server)
```

## Injecting dependencies

Where are our dependencies specified at present?

- custom and third-party scripts are in `./inst/assets/`
- `htmlDependency` for these is defined in `boxxyOutput()`

`countup.js` is required for the animation ... and only for the animation.

Imagine you don't _always_ want to use animation.

Easy:

  - change `boxxy(..., animate = FALSE)`
  - put a switch in `renderValue()` that prevents animation if `data.animate` is `false`

But:

  - `countup.js` is not required if we aren't animating the numbers
  - neither `boxxy()` nor `renderValue()` are used in `boxxyOutput`
  - so we would need a second user-specified `animate = FALSE` argument in `boxxyOutput()` to
  control the dependencies.
  
An alternative:

  - specify optional dependencies in the `renderBoxxy()` function

Workflow:

- optionally ...
- define an `htmlDependency()` in your `render<OutputName>()` function
- serialise that `htmlDependency()` using `shiny::createWebDependency()`
- run `Shiny.renderDependencies()` in the browser to attach the dependency
- (and tweak the logic of your `renderValue()` method)

With this, the dependency is not transferred unless it is required.

Note that we used `htmltools::attachDependencies()` to specify the fixed dependencies in
`boxxyOutput()`

## Summary

With a custom output you have to:

- create the data you want to use (`boxxy()`)
- serialise that data and transmit it to the front end (`renderBoxxy()`)
- construct the HTML skeleton within which the data/images will be presented (`boxxyOutput`)
- tell front-end shiny what to do with the data you have sent it
  - create an OutputBinding (one for each type of custom output)
  - add a `find` method to the OutputBinding (how to find output elements _of this type_)
  - add a `renderValue` method to the OutputBinding (how to update an output element in response to
  new data)
  - register the OutputBinding with front-end shiny
- and also transfer any custom or third-party scripts that are required for the above

## Meeting Videos

### Cohort 1

Week 1 video:

`r knitr::include_url("https://www.youtube.com/embed/pPau7AzQAww")`

<details>
<summary> Meeting chat log </summary>

```
Week 1
00:47:06	Ryan Metcalf:	Absolute versus Relative paths: https://www.computerhope.com/issues/ch001708.htm
01:01:55	LUCIO ENRIQUE CORNEJO RAM REZ:	I have an idea why the author did not use shiny::createRenderFunction
01:02:42	Arthur Shaw:	@lucio, please jump in! I'm just as confused as Russ.
01:12:26	LUCIO ENRIQUE CORNEJO RAM REZ:	I hope i get sick again so I can come to this sessions, instead of going to class :)

Week 2
00:14:24	Info Training:	https://shiny.rstudio.com/articles/building-outputs.html
00:22:29	Ryan Metcalf:	I apologize team. I must drop. I'll catch the recording post production. Multitasking on my end ;). Have a great day!
00:22:38	Info Training:	No problem
```
</details>
